---
title: "Msc Financial Risk Management Diss - Long Memory GARCH (FIGARCH/FIAPARCH) Master"
author: "Demi Gray"
date: "2026-08-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 4)
```

## 1. Packages

```{r package-imports, message=FALSE, warning=FALSE}
library(quantmod)
library(xts)
library(zoo)
library(dplyr)
library(ggplot2)
library(fEGarch)
library(evir)
library(POT)
library(esback)
library(moments)
library(tseries)
library(urca)
```

## 2. CSV Import & Log Return Construction

```{r data-import}
wti_raw   <- read.csv("WTI SPOT CSV.csv",   skip = 5, header = TRUE)
brent_raw <- read.csv("BRENT SPOT CSV.csv", skip = 5, header = TRUE)

colnames(wti_raw)   <- c("Date", "Price")
colnames(brent_raw) <- c("Date", "Price")

wti_raw$Date   <- as.Date(wti_raw$Date,   format = "%m/%d/%Y")
brent_raw$Date <- as.Date(brent_raw$Date, format = "%m/%d/%Y")

wti_raw   <- wti_raw[!is.na(wti_raw$Date)   & !is.na(wti_raw$Price), ]
brent_raw <- brent_raw[!is.na(brent_raw$Date) & !is.na(brent_raw$Price), ]

wti_raw   <- wti_raw[order(wti_raw$Date), ]
brent_raw <- brent_raw[order(brent_raw$Date), ]

wti_xts   <- xts(wti_raw$Price,   order.by = wti_raw$Date)
brent_xts <- xts(brent_raw$Price, order.by = brent_raw$Date)
colnames(wti_xts) <- "Price"; colnames(brent_xts) <- "Price"
```

```{r trim-and-clean}
start_date <- as.Date("2010-01-01")
end_date   <- as.Date("2025-12-31")

wti_xts   <- wti_xts[paste(start_date, end_date, sep = "/")]
brent_xts <- brent_xts[paste(start_date, end_date, sep = "/")]

# Exclude the 20 April 2020 negative WTI price (log-return undefined for P <= 0)
wti_xts <- wti_xts[index(wti_xts) != as.Date("2020-04-20"), ]
```

```{r construct-returns}
wti_ret   <- 100 * diff(log(wti_xts))
brent_ret <- 100 * diff(log(brent_xts))
wti_ret   <- na.omit(wti_ret)
brent_ret <- na.omit(brent_ret)

# Exclude the contaminated 21 April 2020 return (downstream of the excluded 20 April price)
wti_ret <- wti_ret[index(wti_ret) != as.Date("2020-04-21"), ]

cat("WTI observations:",   nrow(wti_ret),   "\n")
cat("Brent observations:", nrow(brent_ret), "\n")
```

## 3. Preliminary Analysis

Identical to the short-memory suite, since these results depend only on
the return series itself, not on the choice of GARCH specification. The
full preliminary battery (stationarity, visual inspection, mean-equation
specification check, zero-mean t-test) is conducted once in the
short-memory master file; its conclusions are carried forward here.

```{r descriptive-stats}
desc_stats <- function(x, name) {
  x <- as.numeric(x)
  jb <- jarque.bera.test(x)
  data.frame(Series = name, Mean = mean(x), SD = sd(x),
             Skewness = skewness(x), ExcKurt = kurtosis(x) - 3,
             Min = min(x), Max = max(x),
             JB_stat = unname(jb$statistic), JB_pvalue = jb$p.value)
}
rbind(desc_stats(wti_ret, "WTI"), desc_stats(brent_ret, "Brent"))
```

```{r ljung-box-prelim}
lb_row <- function(x, lag, name) {
  x <- as.numeric(x)
  raw <- Box.test(x, lag = lag, type = "Ljung-Box")
  sq  <- Box.test(x^2, lag = lag, type = "Ljung-Box")
  data.frame(Series = name, Lag = lag, LB_raw_p = raw$p.value, LB_sq_p = sq$p.value)
}
lags <- c(5, 10, 20)
do.call(rbind, c(lapply(lags, lb_row, x = wti_ret, name = "WTI"),
                lapply(lags, lb_row, x = brent_ret, name = "Brent")))
```

Conclusion carried forward from the short-memory suite: a zero-mean
specification is adequate for both series (Ljung-Box p = 0.76 WTI,
p = 0.30 Brent at lag 10 on standardised residuals; AR(1) inclusion did
not materially improve fit). Zero-mean is enforced throughout below via
`mean_spec(orders = c(0,0), include_mean = FALSE)`.

## 4. Model Estimation: FIGARCH & FIAPARCH (Norm + Student-t, Zero-Mean)

The 8-model configuration list defined here drives everything
downstream (gate, EVT, rolling), so a model is specified in exactly one
place.

```{r figarch-fiaparch-fit}
zero_mean <- mean_spec(orders = c(0, 0), include_mean = FALSE)

# Master configuration: name -> return series, fitting function, distribution
lm_configs <- list(
  WTI_FIGARCH_norm    = list(ret = wti_ret,   fun = figarch,  dist = "norm"),
  WTI_FIGARCH_std     = list(ret = wti_ret,   fun = figarch,  dist = "std"),
  WTI_FIAPARCH_norm   = list(ret = wti_ret,   fun = fiaparch, dist = "norm"),
  WTI_FIAPARCH_std    = list(ret = wti_ret,   fun = fiaparch, dist = "std"),
  Brent_FIGARCH_norm  = list(ret = brent_ret, fun = figarch,  dist = "norm"),
  Brent_FIGARCH_std   = list(ret = brent_ret, fun = figarch,  dist = "std"),
  Brent_FIAPARCH_norm = list(ret = brent_ret, fun = fiaparch, dist = "norm"),
  Brent_FIAPARCH_std  = list(ret = brent_ret, fun = fiaparch, dist = "std")
)

lm_models <- lapply(lm_configs, function(cfg) {
  cfg$fun(as.numeric(cfg$ret), orders = c(1, 1), cond_dist = cfg$dist,
          meanspec = zero_mean)
})
```

```{r inspect-all-fits}
for (nm in names(lm_models)) {
  cat("\n===", nm, "===\n")
  print(lm_models[[nm]])
}
```

### 4.1 Convergence / Estimation Sanity Check

Mirrors the short-memory suite's convergence check (Section 5.1b there).
fEGarch does not expose a `convergence()` flag like rugarch, so the
check is on the estimation output itself: the log-likelihood must be
finite and no coefficient standard error may be NA/NaN (a failed or
degenerate Rsolnp solution typically produces a non-invertible Hessian
and hence NA standard errors). The fEGarch authors' own simulation study
(Schulz et al. 2026, Table 6) reports convergence issues arising
specifically for long-memory specifications, which motivates checking
explicitly here.

```{r convergence-check-lm}
conv_check_lm <- function(fit, name) {
  ll  <- tryCatch(as.numeric(fit@llhood), error = function(e) NA_real_)
  ses <- tryCatch(fit@se, error = function(e) NA_real_)
  data.frame(Model = name,
             LogLik_finite = is.finite(ll),
             All_SE_valid  = all(is.finite(ses)),
             OK = is.finite(ll) && all(is.finite(ses)))
}
conv_table_lm <- do.call(rbind, Map(conv_check_lm, lm_models, names(lm_models)))
conv_table_lm

if (!all(conv_table_lm$OK)) {
  cat("WARNING: the following fits show signs of estimation failure and",
      "should be investigated before proceeding:\n")
  print(conv_table_lm[!conv_table_lm$OK, ])
} else {
  cat("All 8 fits show finite log-likelihoods and valid standard errors.\n")
}
```

### 4.2 Residual-Type Sanity Check

The diagnostic gate and the entire EVT layer operate on standardised
residuals (eta_t = r_t / sigma_t). This one-off check confirms
`residuals()` on an fEGarch fit returns the standardised series (@etat)
rather than raw residuals, so nothing downstream is silently fitted to
the wrong series.

```{r residual-sanity-lm}
f_chk <- lm_models[[1]]
chk1 <- isTRUE(all.equal(as.numeric(residuals(f_chk)), as.numeric(f_chk@etat)))
chk2 <- isTRUE(all.equal(as.numeric(residuals(f_chk)),
                         as.numeric(f_chk@rt) / as.numeric(sigt(f_chk))))
cat("residuals() == @etat:", chk1, "\n")
cat("residuals() == rt/sigt:", chk2, "\n")
if (!chk1 && !chk2) stop("residuals() does not return standardised residuals - use fit@etat downstream instead.")
```

### 4.3 Non-Negativity Diagnostic (Baillie et al. 1996)

Fitted parameters do not satisfy the Baillie, Bollerslev & Mikkelsen
(1996) sufficient (not necessary) restriction phi <= d <= 1-beta for
guaranteed non-negative conditional variance. Since this condition is
sufficient rather than necessary, the true (and considerably more
complex) region of parameter space guaranteeing non-negativity is larger
than this convenient box constraint -- a fitted model can therefore lie
outside the box while still never producing a negative variance. This is
checked directly below via the fitted conditional variance path itself,
rather than relying on the simplified sufficient condition alone.

```{r negativity-check}
check_negativity <- function(fit, name) {
  sig <- tryCatch(sigt(fit), error = function(e) NULL)
  if (is.null(sig)) return(data.frame(Model = name, N_negative = NA, Min_value = NA))
  data.frame(Model = name, N_negative = sum(sig < 0), Min_value = round(min(sig), 6))
}
do.call(rbind, Map(check_negativity, lm_models, names(lm_models)))
```

Direct inspection confirms conditional variance stayed strictly positive
throughout the full sample for all 8 models, despite the sufficient
condition not being met.

## 5. Diagnostic Gate

Same explicit pass/fail rule as the short-memory suite: standardised
residuals must show no significant autocorrelation (Ljung-Box, raw and
squared, p > 0.05 at lag 10). All models passing are carried forward to
EVT, the rolling engine, and backtesting -- the number of survivors is
determined by the data, not fixed in advance.

```{r diagnostic-gate-lm}
diag_check_lm <- function(fit, name) {
  z <- as.numeric(residuals(fit))
  lb_raw <- Box.test(z,   lag = 10, type = "Ljung-Box")
  lb_sq  <- Box.test(z^2, lag = 10, type = "Ljung-Box")
  data.frame(Model = name, LB_raw_p = round(lb_raw$p.value, 4), LB_sq_p = round(lb_sq$p.value, 4))
}

lm_diag_table <- do.call(rbind, Map(diag_check_lm, lm_models, names(lm_models)))
lm_diag_table

gate_pass_lm <- lm_diag_table$Model[lm_diag_table$LB_raw_p > 0.05 & lm_diag_table$LB_sq_p > 0.05]
cat("\nModels passing the diagnostic gate:\n"); print(gate_pass_lm)

lm_models_survivors  <- lm_models[gate_pass_lm]
lm_configs_survivors <- lm_configs[gate_pass_lm]
```

## 6. Model Selection (AIC/BIC)

Selection here names the primary specification for headline discussion
only; all diagnostically adequate models are retained through rolling
backtesting, with model discrimination left to out-of-sample
performance (consistent with the short-memory suite).

```{r ic-table-lm}
# fEGarch objects are S4 and do not carry logLik()/AIC()/BIC() methods.
# Per Schulz et al. (2026): @llhood is the (unnormalised) log-likelihood,
# @inf_criteria holds the normalised AIC and BIC (AIC/n, BIC/n). n is
# identical within each series, so model rankings are unaffected.
ic_row_lm <- function(fit, name) {
  ic <- fit@inf_criteria
  data.frame(Model = name,
             LogLik = round(as.numeric(fit@llhood), 2),
             AIC_n  = round(unname(ic[1]), 4),
             BIC_n  = round(unname(ic[2]), 4))
}
do.call(rbind, Map(ic_row_lm, lm_models, names(lm_models)))
```

FIAPARCH-std is the primary specification for both series (lowest
AIC/BIC among survivors), consistent with Aloui & Mabrouk (2010).
FIGARCH consistently estimates d near the unit-root boundary
(0.95-1.00); FIAPARCH gives more moderate d (~0.53-0.58) once
asymmetry/power flexibility is added.

### 6.1 Fractional Differencing & Distribution Shape Parameters

Mirrors the short-memory suite's degrees-of-freedom table (5.3b there),
extended with the long-memory parameter d. Extraction is by name pattern
with the full coefficient vector printed once, so a naming mismatch is
visible rather than silently returning NA.

```{r param-table-lm}
cat("Coefficient names (first model):",
    paste(names(lm_models[[1]]@pars), collapse = ", "), "\n\n")

param_row_lm <- function(fit, name) {
  cf <- fit@pars; se <- tryCatch(fit@se, error = function(e) rep(NA, length(cf)))
  pick <- function(pat) {
    i <- grep(pat, names(cf))
    if (length(i) == 0) c(NA, NA) else c(unname(cf[i[1]]), unname(se[i[1]]))
  }
  d_v  <- pick("^d$")
  sh_v <- pick("^df$")
  data.frame(Model = name,
             d = round(d_v[1], 4), d_SE = round(d_v[2], 4),
             Shape = round(sh_v[1], 4), Shape_SE = round(sh_v[2], 4))
}
do.call(rbind, Map(param_row_lm, lm_models, names(lm_models)))
```

### 6.2 In-Sample Forecast Evaluation (MSE, MAE, QLIKE)

Mirrors the short-memory suite's in-sample evaluation (5.4 there):
fitted conditional variance against squared returns over the full
sample, complementing the out-of-sample metrics in Section 13.

```{r insample-accuracy-lm}
insample_accuracy_lm <- function(fit, cfg, name) {
  realized_var <- as.numeric(cfg$ret)^2
  fitted_var   <- as.numeric(sigt(fit))^2
  data.frame(Model = name,
             MSE  = round(mean((realized_var - fitted_var)^2, na.rm = TRUE), 4),
             MAE  = round(mean(abs(realized_var - fitted_var), na.rm = TRUE), 4),
             QLIKE = round(mean(log(fitted_var) + realized_var / fitted_var, na.rm = TRUE), 4))
}
do.call(rbind, Map(insample_accuracy_lm, lm_models, lm_configs, names(lm_models)))
```

## 7. News Impact Surface (FIAPARCH Models)

# --- News Impact Surface for FIAPARCH models ---
# FIAPARCH's long-memory structure means the true "impact" of a single
# shock formally depends on an infinite ARCH(∞) representation via the
# fractional-differencing filter. Following standard practice (e.g. Tse
# 1998), this is approximated by truncating that representation and
# examining the leading-order response: how conditional variance reacts
# to today's shock, holding the lagged variance term at fixed multiples
# of the unconditional variance (mirroring the MS-GARCH NIS construction).
```{r NIC}
fiaparch_nis <- function(fit, z_range = seq(-4, 4, length.out = 60),
                          lagvar_mult = seq(0.5, 2.5, length.out = 40)) {
  pars <- fit@pars
  omega <- pars["omega"]
  beta1 <- pars["beta1"]
  gamma <- pars["gamma"]
  delta <- pars["delta"]

  uncond_var <- (omega / (1 - beta1))^(2/delta)  # approx unconditional variance scale

  grid <- expand.grid(z = z_range, lagvar_ratio = lagvar_mult)

  grid$sigma_lag_delta <- (uncond_var * grid$lagvar_ratio)^(delta/2)

  # Asymmetric power transformation (APARCH-style): (|z| - gamma*z)^delta
  grid$shock_term <- (abs(grid$z) - gamma * grid$z)^delta

  # Leading-order approximation: variance response combines the shock
  # term with the persisted lagged-variance term, scaled by omega
  grid$cond_var <- (omega + grid$shock_term + beta1 * grid$sigma_lag_delta)^(2/delta)

  grid
}

nis_wti_fiaparch   <- fiaparch_nis(lm_models[["WTI_FIAPARCH_std"]])
nis_brent_fiaparch <- fiaparch_nis(lm_models[["Brent_FIAPARCH_std"]])

library(ggplot2)
ggplot(nis_wti_fiaparch, aes(x = z, y = lagvar_ratio, fill = cond_var)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Conditional\nVariance") +
  labs(title = "WTI FIAPARCH-std: News Impact Surface",
       x = "Standardized shock (z)", y = "Lagged variance (× unconditional)") +
  theme_minimal()

ggplot(nis_brent_fiaparch, aes(x = z, y = lagvar_ratio, fill = cond_var)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Conditional\nVariance") +
  labs(title = "Brent FIAPARCH-std: News Impact Surface",
       x = "Standardized shock (z)", y = "Lagged variance (× unconditional)") +
  theme_minimal()
```

## 8. EVT: Standardised Residuals, MRL Plots, GPD Fits (u=90%)

```{r evt-residuals-lm}
z_lm <- lapply(lm_models_survivors, function(fit) as.numeric(residuals(fit)))
```

```{r evt-mrl-lm}
for (m in names(z_lm)) {
  par(mfrow = c(1, 2))
  mrlplot(z_lm[[m]],  main = paste("Right Tail MRL -", m))
  abline(v = quantile(z_lm[[m]], 0.90), col = "red", lty = 2)
  mrlplot(-z_lm[[m]], main = paste("Left Tail MRL -", m))
  abline(v = quantile(-z_lm[[m]], 0.90), col = "red", lty = 2)
  par(mfrow = c(1, 1))
}
```

```{r evt-gpd-lm}
fit_gpd_row_lm <- function(x, name, tail, q = 0.90) {
  u <- unname(quantile(x, q))
  fit <- fitgpd(x, threshold = u, est = "mle")
  data.frame(Model = name, Tail = tail, Percentile = q * 100, U = round(u, 4),
             Shape = round(unname(fit$fitted.values["shape"]), 4),
             Shape_SE = round(unname(fit$std.err["shape"]), 4),
             Scale = round(unname(fit$fitted.values["scale"]), 4),
             Scale_SE = round(unname(fit$std.err["scale"]), 4),
             N_u = fit$nat)
}

gpd_results_lm <- do.call(rbind, lapply(names(z_lm), function(m) {
  rbind(fit_gpd_row_lm(z_lm[[m]],  m, "Right"),
        fit_gpd_row_lm(-z_lm[[m]], m, "Left"))
}))
gpd_results_lm
saveRDS(gpd_results_lm, "longmemory_gpd_shape_table.rds")
```

## 9. Rolling Out-of-Sample Engine (fEGarch Native: steady_window=TRUE)

Uses fEGarch's native `predict_roll()` with `steady_window = TRUE` for a
genuine fixed-width moving window,
matching the 1,000-observation window and 20-day refit cadence used in
the short-memory and MS-GARCH suites. `n_test` is set so the initial
training window is exactly 1,000 observations.

Consistent with the short-memory suite, ALL diagnostic-gate survivors
are rolled, not only the AIC/BIC winner. GPD tails are refit at the same
20-day cadence. Where fEGarch's refit-enabled `predict_roll()` output
exposes the per-refit fitted models (`@model` as a list; Schulz et al.
2026, Section 5.3), residuals for the GPD refits are taken directly from
those objects, so the tail estimates come from the identical fits that
produce the sigma forecasts (matching the short-memory engine's
internal consistency) at no extra estimation cost. If that slot is not a
usable list, the engine falls back to manually refitting on the same
1,000-day windows at the same boundaries (equivalent fits, roughly
double the run time).

Both ES confidence levels (97.5% and 99%) are computed in the same pass:
VaR99, ES97.5 and ES99 are all quantile-based functionals of the same
fitted GPD tail on the same sigma path, so a second rolling run per ES
level (as in the short-memory file, where tail parameters were not
retained per step) is redundant -- the resulting numbers are identical.

```{r rolling-function-lm}
gpd_tail_quantiles <- function(x, q_thresh = 0.90, p_var = 0.99,
                               p_es_a = 0.975, p_es_b = 0.99) {
  u  <- quantile(x, q_thresh)
  gf <- fitgpd(x, threshold = u, est = "mle")
  xi   <- unname(gf$fitted.values["shape"])
  beta <- unname(gf$fitted.values["scale"])
  n    <- length(x); nu_ex <- sum(x > u)
  q_gpd <- function(p) u + (beta / xi) * (((1 - p) / (nu_ex / n))^(-xi) - 1)
  es_of <- function(qp) qp / (1 - xi) + (beta - xi * u) / (1 - xi)
  list(z_var    = q_gpd(p_var),
       z_es_a   = es_of(q_gpd(p_es_a)),
       z_es_b   = es_of(q_gpd(p_es_b)))
}

run_longmemory_evt_roll <- function(ret_xts, model_fun, window_size = 1000,
                                    refit_every = 20, q_thresh = 0.90,
                                    p_var = 0.99, p_es_a = 0.975, p_es_b = 0.99,
                                    cond_dist = "std", label = "") {
  y_all <- as.numeric(ret_xts)
  n     <- length(y_all)
  n_test <- n - window_size

  zero_mean <- mean_spec(orders = c(0, 0), include_mean = FALSE)

  fit0 <- tryCatch(
    model_fun(y_all, orders = c(1,1), cond_dist = cond_dist,
             meanspec = zero_mean, n_test = n_test),
    error = function(e) NULL
  )
  if (is.null(fit0)) stop("Initial fit with n_test failed for ", label)

  roll <- predict_roll(fit0, step_size = 1, refit_after = refit_every,
                       steady_window = TRUE)

  sig_roll <- as.numeric(sigt(roll))
  test_obs <- y_all[(window_size + 1):n]
  dates    <- index(ret_xts)[(window_size + 1):n]

  n_fc <- length(sig_roll)
  n_refits <- ceiling(n_fc / refit_every)

  # Try to reuse the per-refit fitted models from predict_roll() itself
  # (Schulz et al. 2026, Sec. 5.3: @model is a list under refitting).
  refit_models <- tryCatch({
    ml <- roll@model
    if (is.list(ml) && length(ml) >= n_refits) ml else NULL
  }, error = function(e) NULL)
  use_native_refits <- !is.null(refit_models)
  cat(label, "- refit residual source:",
      if (use_native_refits) "native (roll@model)" else "manual fallback", "\n")

  out <- data.frame(date = dates[seq_len(n_fc)], realized = test_obs[seq_len(n_fc)],
                    sigma = sig_roll,
                    VaR99_evt_L = NA_real_, ES975_evt_L = NA_real_, ES99_evt_L = NA_real_,
                    VaR99_evt_R = NA_real_, ES975_evt_R = NA_real_, ES99_evt_R = NA_real_)

  tail_L <- NULL; tail_R <- NULL
  for (i in seq_len(n_fc)) {
    if ((i - 1) %% refit_every == 0) {
      k <- (i - 1) %/% refit_every + 1
      z_in <- NULL
      if (use_native_refits) {
        z_in <- tryCatch(as.numeric(residuals(refit_models[[k]])), error = function(e) NULL)
      }
      if (is.null(z_in)) {
        hist_end <- window_size + i - 1
        w <- y_all[max(1, hist_end - window_size + 1):hist_end]
        f_i <- tryCatch(model_fun(w, orders = c(1,1), cond_dist = cond_dist,
                                  meanspec = zero_mean),
                        error = function(e) NULL)
        if (!is.null(f_i)) z_in <- as.numeric(residuals(f_i))
      }
      if (!is.null(z_in)) {
        tail_L_new <- tryCatch(gpd_tail_quantiles(-z_in, q_thresh, p_var, p_es_a, p_es_b), error = function(e) NULL)
        tail_R_new <- tryCatch(gpd_tail_quantiles( z_in, q_thresh, p_var, p_es_a, p_es_b), error = function(e) NULL)
        if (!is.null(tail_L_new)) tail_L <- tail_L_new
        if (!is.null(tail_R_new)) tail_R <- tail_R_new
        # else: hold previous window's tail fit (fallback, matching short-memory engine)
      }
      cat(label, "refit", k, "of", n_refits, "(forecast step", i, "of", n_fc, ")\n")
    }
    if (is.null(tail_L) || is.null(tail_R)) next
    sig_i <- sig_roll[i]
    out$VaR99_evt_L[i] <- -sig_i * tail_L$z_var
    out$ES975_evt_L[i] <- -sig_i * tail_L$z_es_a
    out$ES99_evt_L[i]  <- -sig_i * tail_L$z_es_b
    out$VaR99_evt_R[i] <-  sig_i * tail_R$z_var
    out$ES975_evt_R[i] <-  sig_i * tail_R$z_es_a
    out$ES99_evt_R[i]  <-  sig_i * tail_R$z_es_b
  }
  out
}
```

### 9.1 Rolling Run: All Survivors (u=90%, Refit-20, Both ES Levels)

Each model's result is saved to its own RDS immediately on completion,
so a failure partway through the loop loses at most the model currently
running (partial substitute for the checkpointing built into the
short-memory engine, which is not possible inside native
`predict_roll()`).

```{r run-rolling-lm, eval=FALSE}
rolling_results_lm <- list()

for (m in names(lm_configs_survivors)) {
  cfg <- lm_configs_survivors[[m]]
  save_file <- paste0("lm_roll_", m, ".rds")
  if (file.exists(save_file)) {
    cat("===", m, ": already done, loading from", save_file, "===\n")
    rolling_results_lm[[m]] <- readRDS(save_file)
    next
  }
  cat("===", m, "(refit-20, u=90%) ===\n")
  res <- tryCatch(
    run_longmemory_evt_roll(cfg$ret, cfg$fun, cond_dist = cfg$dist, label = m),
    error = function(e) { cat("FAILED:", m, "-", conditionMessage(e), "\n"); NULL }
  )
  if (!is.null(res)) {
    rolling_results_lm[[m]] <- res
    saveRDS(res, save_file)
  }
}

saveRDS(rolling_results_lm, "longmemory_evt_roll_all.rds")
```

```{r violations-lm, eval=FALSE}
violation_summary <- function(roll, label) {
  viol_L <- roll$realized < roll$VaR99_evt_L
  viol_R <- roll$realized > roll$VaR99_evt_R
  cat(label, "- Left:", sum(viol_L, na.rm = TRUE),
      "(", round(100*mean(viol_L, na.rm = TRUE), 3), "%) | Right:",
      sum(viol_R, na.rm = TRUE), "(", round(100*mean(viol_R, na.rm = TRUE), 3), "%)\n")
}
for (m in names(rolling_results_lm)) violation_summary(rolling_results_lm[[m]], m)
```

## 10. Backtest Functions (esback::er_backtest, seeded)

```{r backtest-functions-lm}
kupiec_test <- function(violations, alpha) {
  n <- length(violations); x <- sum(violations, na.rm = TRUE); pi_hat <- x / n
  if (x == 0 || x == n) return(list(LR = NA, p = NA, rate = pi_hat, n_viol = x))
  LR <- -2 * (log((1 - alpha)^(n - x) * alpha^x) - log((1 - pi_hat)^(n - x) * pi_hat^x))
  list(LR = LR, p = 1 - pchisq(LR, df = 1), rate = pi_hat, n_viol = x)
}

christoffersen_test <- function(violations) {
  v <- as.integer(violations)
  n00 <- sum(v[-length(v)] == 0 & v[-1] == 0); n01 <- sum(v[-length(v)] == 0 & v[-1] == 1)
  n10 <- sum(v[-length(v)] == 1 & v[-1] == 0); n11 <- sum(v[-length(v)] == 1 & v[-1] == 1)
  p01 <- n01 / (n00 + n01); p11 <- n11 / (n10 + n11)
  p   <- (n01 + n11) / (n00 + n01 + n10 + n11)
  if (any(is.nan(c(p01, p11, p)))) return(list(LR_ind = NA, p_ind = NA))
  ll_uncond <- function(pp) (n00 + n10) * log(1 - pp) + (n01 + n11) * log(pp)
  ll_ind <- (n00*log(1-p01) + n01*log(p01) + n10*log(1-p11) + n11*log(p11))
  LR_ind <- -2 * (ll_uncond(p) - ll_ind)
  list(LR_ind = LR_ind, p_ind = 1 - pchisq(LR_ind, df = 1))
}

esback_test <- function(realized, VaR, ES, tail = c("left", "right"), seed = 42, B = 1000) {
  tail <- match.arg(tail)
  set.seed(seed)
  ok <- !is.na(VaR) & !is.na(ES)
  realized <- realized[ok]; VaR <- VaR[ok]; ES <- ES[ok]
  if (tail == "left") {
    violated <- realized < VaR
    r <- realized[violated]; q <- VaR[violated]; e <- ES[violated]
  } else {
    violated <- realized > VaR
    r <- -realized[violated]; q <- -VaR[violated]; e <- -ES[violated]
  }
  n_viol <- sum(violated)
  if (n_viol < 5) return(list(n_viol = n_viol, p_twosided = NA))
  result <- tryCatch(esback::er_backtest(r = r, q = q, e = e, B = B), error = function(e) NULL)
  if (is.null(result)) return(list(n_viol = n_viol, p_twosided = NA))
  list(n_viol = n_viol, p_twosided = round(result$pvalue_twosided_simple, 4))
}

runs_test <- function(violations) {
  v <- as.integer(violations); n <- length(v); n1 <- sum(v, na.rm = TRUE); n0 <- n - n1
  if (n1 == 0 || n0 == 0) return(list(z = NA, p = NA))
  runs <- 1 + sum(diff(v) != 0, na.rm = TRUE)
  mu   <- 2 * n1 * n0 / n + 1
  var_ <- 2 * n1 * n0 * (2 * n1 * n0 - n) / (n^2 * (n - 1))
  z <- (runs - mu) / sqrt(var_)
  list(z = z, p = 2 * (1 - pnorm(abs(z))))
}
```

## 11. Primary Backtest (All Survivors: u=90%, Refit-20, VaR99/ES97.5)

```{r backtest-run-lm, eval=FALSE}
run_backtest_row_lm <- function(roll, tail = c("left", "right"), label = "",
                                es_col_suffix = "ES975_evt") {
  tail <- match.arg(tail)
  tsuf <- if (tail == "left") "_L" else "_R"
  VaR <- roll[[paste0("VaR99_evt", tsuf)]]
  ES  <- roll[[paste0(es_col_suffix, tsuf)]]
  realized <- roll$realized
  viol <- if (tail == "left") realized < VaR else realized > VaR

  kup <- kupiec_test(viol[!is.na(viol)], 0.01)
  chr <- christoffersen_test(viol[!is.na(viol)])
  eb  <- esback_test(realized, VaR, ES, tail)
  run <- runs_test(viol[!is.na(viol)])

  data.frame(Model = label, Tail = tail, n_viol = kup$n_viol,
             rate_pct = round(100 * kup$rate, 3), Kupiec_p = round(kup$p, 4),
             Christoffersen_p = round(chr$p_ind, 4),
             esback_p = eb$p_twosided, Runs_p = round(run$p, 4))
}

run_all_backtests_lm <- function(roll_list, es_col_suffix = "ES975_evt") {
  do.call(rbind, unlist(lapply(names(roll_list), function(m) {
    list(run_backtest_row_lm(roll_list[[m]], "left",  m, es_col_suffix),
         run_backtest_row_lm(roll_list[[m]], "right", m, es_col_suffix))
  }), recursive = FALSE))
}

backtest_lm_primary <- run_all_backtests_lm(rolling_results_lm, "ES975_evt")
backtest_lm_primary
write.csv(backtest_lm_primary, "longmemory_backtest_primary_esback.csv", row.names = FALSE)
```

## 12. VaR99/ES97.5 vs VaR99/ES99 Comparison (u=90%, Refit-20 Fixed)

No threshold sweep is performed; the GPD threshold is fixed at the
MRL-confirmed 90th percentile throughout, matching the short-memory
suite. Two confidence pairings are compared at this fixed threshold: the
primary Basel FRTB pairing (VaR99/ES97.5) and a stricter pairing
(VaR99/ES99), to test whether the ES confidence level itself affects
exceedance-residual calibration. Both were computed in the single
rolling pass above (identical sigma path and GPD fits; only the ES
quantile differs), so no second run is required.

```{r es-comparison-backtest-lm, eval=FALSE}
es_comparison_lm <- rbind(
  cbind(ES_level = "97.5%", run_all_backtests_lm(rolling_results_lm, "ES975_evt")),
  cbind(ES_level = "99%",   run_all_backtests_lm(rolling_results_lm, "ES99_evt"))
)
es_comparison_lm
write.csv(es_comparison_lm, "longmemory_es_comparison.csv", row.names = FALSE)
```

## 13. Out-of-Sample Forecast Accuracy (MSE, MAE, QLIKE)

Matches the metrics used in the short-memory and MS-GARCH suites for
consistency in the eventual cross-framework comparison.

```{r oos-accuracy-lm, eval=FALSE}
oos_accuracy_lm <- function(roll, label) {
  realized_var <- roll$realized^2
  forecast_var <- roll$sigma^2
  data.frame(Model = label,
             MSE  = round(mean((realized_var - forecast_var)^2, na.rm = TRUE), 4),
             MAE  = round(mean(abs(realized_var - forecast_var), na.rm = TRUE), 4),
             QLIKE = round(mean(log(forecast_var) + realized_var / forecast_var, na.rm = TRUE), 4))
}

oos_lm_table <- do.call(rbind, lapply(names(rolling_results_lm), function(m) {
  oos_accuracy_lm(rolling_results_lm[[m]], m)
}))
oos_lm_table
write.csv(oos_lm_table, "longmemory_oos_accuracy.csv", row.names = FALSE)
```
```{r save-final-dminput, eval=FALSE}
saveRDS(list(
  wti   = list(realized = longmemory_evt_roll_all[["WTI_FIAPARCH_std"]]$realized,
               VaR_L    = longmemory_evt_roll_all[["WTI_FIAPARCH_std"]]$VaR99_evt_L,
               VaR_R    = longmemory_evt_roll_all[["WTI_FIAPARCH_std"]]$VaR99_evt_R,
               Date     = longmemory_evt_roll_all[["WTI_FIAPARCH_std"]]$date),
  brent = list(realized = longmemory_evt_roll_all[["Brent_FIAPARCH_std"]]$realized,
               VaR_L    = longmemory_evt_roll_all[["Brent_FIAPARCH_std"]]$VaR99_evt_L,
               VaR_R    = longmemory_evt_roll_all[["Brent_FIAPARCH_std"]]$VaR99_evt_R,
               Date     = longmemory_evt_roll_all[["Brent_FIAPARCH_std"]]$date)
), file = "longmemory_dm_input.rds")

cat("Long-memory DM input saved.\n")
```

## 14. Save Final Outputs (Consistent Naming for Cross-Framework Comparison)

```{r save-final-lm, eval=FALSE}
saveRDS(list(
  fits = lm_models,
  gpd_table = gpd_results_lm,
  rolling_results = rolling_results_lm,
  backtest_primary = backtest_lm_primary,
  es_comparison = es_comparison_lm,
  oos_accuracy = oos_lm_table
), file = "longmemory_final_output.rds")

cat("Long-memory final outputs saved.\n")
```
