---
title: "MS-GARCH-EVT: Full Appendix Master Script"
author: "Demi Gray"
date: "2026-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 4)
```

This is the consolidated, appendix-ready MS-GARCH-EVT script. It contains
every stage of the pipeline actually used and reported: full-sample
multi-start estimation, model
characterisation (selection tables, regime dynamics, plots, news impact
curves), EVT/GPD tail modelling, the rolling out-of-sample engine
(single warm-started fit per refit, with fallback, scoped to the
Student-t models per the finding in Part 2), backtesting via
`esback::er_backtest()` applied to both the primary rolling run and
retrospectively to the already-completed 20-config robustness grid, the
parametric-vs-EVT comparison, and an explicit, quantified limitation
regarding rolling-window optimality (Section 19). Where a stage reuses
results already computed and saved to disk from an earlier session (the
20-config grid, the rolling forecast series itself), this is stated
explicitly rather than re-estimated, to avoid unnecessary recomputation.

# PART 1: FULL-SAMPLE ESTIMATION 

## 1. Packages

```{r package-imports, message=FALSE, warning=FALSE}
library(MSGARCH)
library(xts)
library(zoo)
library(POT)
library(esback)
library(ggplot2)

set.seed(123)
```

## 2. Data Import & Returns Construction

```{r data-import}
wti_raw   <- read.csv("WTI SPOT CSV.csv",   skip = 5, header = TRUE)
brent_raw <- read.csv("BRENT SPOT CSV.csv", skip = 5, header = TRUE)

colnames(wti_raw)   <- c("Date", "Price")
colnames(brent_raw) <- c("Date", "Price")

wti_raw$Date   <- as.Date(wti_raw$Date,   format = "%m/%d/%Y")
brent_raw$Date <- as.Date(brent_raw$Date, format = "%m/%d/%Y")

wti_raw   <- wti_raw[!is.na(wti_raw$Date)   & !is.na(wti_raw$Price), ]
brent_raw <- brent_raw[!is.na(brent_raw$Date) & !is.na(brent_raw$Price), ]

wti_raw   <- wti_raw[order(wti_raw$Date), ]
brent_raw <- brent_raw[order(brent_raw$Date), ]

wti_xts   <- xts(wti_raw$Price,   order.by = wti_raw$Date)
brent_xts <- xts(brent_raw$Price, order.by = brent_raw$Date)
colnames(wti_xts) <- "Price"; colnames(brent_xts) <- "Price"
```

```{r trim-and-clean}
start_date <- as.Date("2010-01-01")
end_date   <- as.Date("2025-12-31")

wti_xts   <- wti_xts[paste(start_date, end_date, sep = "/")]
brent_xts <- brent_xts[paste(start_date, end_date, sep = "/")]

wti_xts <- wti_xts[index(wti_xts) != as.Date("2020-04-20"), ]
```

```{r construct-returns}
wti_ret   <- 100 * diff(log(wti_xts))
brent_ret <- 100 * diff(log(brent_xts))
wti_ret   <- na.omit(wti_ret)
brent_ret <- na.omit(brent_ret)

wti_ret <- wti_ret[index(wti_ret) != as.Date("2020-04-21"), ]

cat("WTI observations:",   nrow(wti_ret),   "\n")
cat("Brent observations:", nrow(brent_ret), "\n")
```

## 3. Model Specifications (Norm + Student-t, K=2)

Shape parameters (nu) are left unconstrained across regimes rather than
tied, so the data can speak to whether tail behaviour itself differs by
regime (see Part 2, Section 8 for the resulting evidence).

```{r specs}
spec_ms2_norm <- CreateSpec(
  variance.spec     = list(model = "sGARCH"),
  distribution.spec = list(distribution = "norm"),
  switch.spec       = list(do.mix = FALSE, K = 2)
)

spec_ms2_std <- CreateSpec(
  variance.spec     = list(model = "sGARCH"),
  distribution.spec = list(distribution = "std"),
  switch.spec       = list(do.mix = FALSE, K = 2)
)

summary(spec_ms2_std)
```

## 4. Multi-Start Estimation Function (50 starts, full sample)

```{r multistart-function}
perturb_par <- function(par, sd_log = 0.25) {
  new <- par * exp(rnorm(length(par), mean = 0, sd = sd_log))
  names(new) <- names(par)
  p_idx <- grepl("^P_", names(new));     new[p_idx] <- pmin(pmax(new[p_idx], 0.02), 0.98)
  b_idx <- grepl("^beta", names(new));   new[b_idx] <- pmin(new[b_idx], 0.995)
  a1_idx <- grepl("^alpha1", names(new)); new[a1_idx] <- pmin(pmax(new[a1_idx], 1e-4), 0.5)
  nu_idx <- grepl("^nu", names(new));    new[nu_idx] <- pmax(new[nu_idx], 2.1)
  new
}

FitML_global <- function(spec, data, n_starts = 50, sd_log = 0.25,
                          seed = 123, verbose = TRUE) {
  set.seed(seed)
  data <- as.numeric(data)
  fits <- list(); lls <- c()

  fit0 <- tryCatch(FitML(spec = spec, data = data, ctr = list(do.se = TRUE)),
                   error = function(e) NULL)
  if (!is.null(fit0)) { fits[["default"]] <- fit0; lls["default"] <- fit0$loglik }
  base_par <- if (!is.null(fit0)) fit0$par else NULL

  for (i in seq_len(n_starts)) {
    par0 <- if (!is.null(base_par)) perturb_par(base_par, sd_log) else NULL
    fit_i <- tryCatch(FitML(spec = spec, data = data,
                            ctr = list(par0 = par0, do.se = TRUE)),
                      error = function(e) NULL)
    if (!is.null(fit_i)) {
      nm <- paste0("start_", i); fits[[nm]] <- fit_i; lls[nm] <- fit_i$loglik
    }
  }
  if (length(fits) == 0) stop("All starts failed to converge.")

  best_nm <- names(which.max(lls)); best_fit <- fits[[best_nm]]
  n_at_max <- sum(abs(lls - max(lls)) < 1e-3)
  if (verbose) {
    cat("Successful starts:", length(lls), "of", n_starts + 1, "\n")
    cat("Best log-likelihood:", round(max(lls), 4), "(from", best_nm, ")\n")
    cat("Starts converging to best LL (tol 1e-3):", n_at_max, "of", length(lls), "\n\n")
  }
  list(fit = best_fit, lls = lls, n_at_max = n_at_max)
}

is_degenerate <- function(par, nu_floor = 2.5) {
  nus <- par[grepl("^nu", names(par))]
  p11 <- par["P_1_1"]; p21 <- par["P_2_1"]
  pi1 <- p21 / (1 - p11 + p21)
  (length(nus) > 0 && any(nus < nu_floor)) || min(pi1, 1 - pi1) < 0.03
}
```

## 5. Estimation: Norm + Student-t, WTI + Brent (50 starts each)

```{r fit-all-models, cache=TRUE}
ms_wti_norm   <- FitML_global(spec_ms2_norm, wti_ret,   n_starts = 50, seed = 223)
ms_wti_std    <- FitML_global(spec_ms2_std,  wti_ret,   n_starts = 50, seed = 224)
ms_brent_norm <- FitML_global(spec_ms2_norm, brent_ret, n_starts = 50, seed = 225)
ms_brent_std  <- FitML_global(spec_ms2_std,  brent_ret, n_starts = 50, seed = 226)
```

### 5.1 Brent-std Degenerate-Mode Screen

The main run's Brent-std candidate is screened for a spurious degenerate
maximum -- a documented failure mode in switching-model estimation where a
regime collapses onto a handful of extreme observations (boundary nu,
near-zero persistence, negligible stable probability) rather than
representing genuine volatility dynamics. An independently-seeded pilot
search provides the replicated alternative if the main candidate is
flagged.

```{r fit-brent-pilot, cache=TRUE}
brent_std_pilot <- FitML_global(spec_ms2_std, brent_ret, n_starts = 25, seed = 126)
```

```{r brent-std-adopt}
if (is_degenerate(ms_brent_std$fit$par)) {
  cat("Brent-std main run flagged as degenerate -- adopting pilot fit.\n")
  ms_brent_std <- brent_std_pilot
} else {
  cat("Brent-std main run passed degeneracy screen.\n")
}
```

```{r all-summaries}
summary(ms_wti_norm$fit)
summary(ms_wti_std$fit)
summary(ms_brent_norm$fit)
summary(ms_brent_std$fit)
```

# PART 2: MODEL CHARACTERISATION

## 7. Model Selection

```{r ic-table}
ic_row <- function(fit, label) {
  data.frame(Model = label, LogLik = round(fit$loglik, 2),
             AIC = round(AIC(fit), 2), BIC = round(BIC(fit), 2))
}

ic_table <- rbind(
  ic_row(ms_wti_norm$fit,   "WTI   MS2-sGARCH-norm"),
  ic_row(ms_wti_std$fit,    "WTI   MS2-sGARCH-std"),
  ic_row(ms_brent_norm$fit, "Brent MS2-sGARCH-norm"),
  ic_row(ms_brent_std$fit,  "Brent MS2-sGARCH-std")
)
ic_table
```

Student-t dominates Gaussian on both AIC and BIC for both series.

```{r k1-vs-k2, cache=TRUE}
spec_k1_norm <- CreateSpec(variance.spec = list(model = "sGARCH"),
                           distribution.spec = list(distribution = "norm"),
                           switch.spec = list(K = 1))
spec_k1_std  <- CreateSpec(variance.spec = list(model = "sGARCH"),
                           distribution.spec = list(distribution = "std"),
                           switch.spec = list(K = 1))

k1_wti_norm   <- FitML(spec_k1_norm, as.numeric(wti_ret))
k1_wti_std    <- FitML(spec_k1_std,  as.numeric(wti_ret))
k1_brent_norm <- FitML(spec_k1_norm, as.numeric(brent_ret))
k1_brent_std  <- FitML(spec_k1_std,  as.numeric(brent_ret))

k_table <- rbind(
  ic_row(k1_wti_norm,       "WTI   K1-sGARCH-norm"),
  ic_row(ms_wti_norm$fit,   "WTI   K2-sGARCH-norm"),
  ic_row(k1_wti_std,        "WTI   K1-sGARCH-std"),
  ic_row(ms_wti_std$fit,    "WTI   K2-sGARCH-std"),
  ic_row(k1_brent_norm,     "Brent K1-sGARCH-norm"),
  ic_row(ms_brent_norm$fit, "Brent K2-sGARCH-norm"),
  ic_row(k1_brent_std,      "Brent K1-sGARCH-std"),
  ic_row(ms_brent_std$fit,  "Brent K2-sGARCH-std")
)
k_table
```

For the std models, AIC favours K=2 for both series; BIC marginally
favours K=1 (BIC's parameter penalty scales with sample size and is
stringent on state count in switching models). K=2 is retained on (i)
AIC, (ii) the strongly persistent, economically coherent regime structure
below, and (iii) out-of-sample literature evidence for MS-GARCH's
advantage (Ardia, Bluteau, Boudt & Catania 2018).

## 8. Regime Characterisation

```{r regime-characterisation}
regime_vols <- function(fit_obj, label) {
  uv <- sqrt(252) * sapply(ExtractStateFit(fit_obj), UncVol)
  cat(label, "- annualised unconditional vol: Regime 1 (low):",
      round(uv[1], 2), "% | Regime 2 (high):", round(uv[2], 2), "%\n")
  invisible(uv)
}

regime_vols(ms_wti_norm$fit,   "WTI  norm")
regime_vols(ms_wti_std$fit,    "WTI  std ")
regime_vols(ms_brent_norm$fit, "Brent norm")
regime_vols(ms_brent_std$fit,  "Brent std ")

TransMat(ms_wti_std$fit)
TransMat(ms_brent_std$fit)
```

**Key finding:** the Gaussian models' regime 2 reaches implausible
annualised vol levels (WTI ~299%, Brent ~208%) with very low persistence
and negligible stable probability -- the signature of an
outlier-absorbing jump state rather than a genuine volatility regime.
Thin-tailed regimes have no mechanism other than a transient variance
spike to accommodate oil's extreme single-day moves, since a Gaussian
distribution assigns vanishingly small probability to such days under
either regime directly. The Student-t models' regimes are economically
sensible (WTI ~23%/~38%; Brent ~31%/~45%) with strong persistence
(expected durations of months to years), consistent with genuine
calm/turbulent epochs. This distortion is why the rolling/EVT/backtest
stages below are scoped to the Student-t models only.

## 9. Smoothed Probabilities & Filtered Volatility

```{r smoothed-plots, fig.height=5}
plot_regimes <- function(fit_obj, ret_xts, label) {
  sp  <- State(fit_obj)$SmoothProb[, 1, 2, drop = TRUE]
  sp  <- sp[seq_len(nrow(ret_xts))]
  vol <- sqrt(252) * Volatility(fit_obj)

  par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
  plot(index(ret_xts), as.numeric(ret_xts), type = "h", col = "grey70",
       main = paste0(label, ": returns & P[high-vol regime]"),
       xlab = "", ylab = "Return (%)")
  par(new = TRUE)
  plot(index(ret_xts), sp, type = "l", col = "red", lwd = 1.5,
       axes = FALSE, xlab = "", ylab = "", ylim = c(0, 1))
  axis(4); mtext("Smoothed prob.", side = 4, line = -1.2, cex = 0.8)

  plot(index(ret_xts), vol[seq_len(nrow(ret_xts))], type = "l",
       main = paste0(label, ": filtered conditional volatility (ann. %)"),
       xlab = "", ylab = "Vol (%)")
  par(mfrow = c(1, 1))
}

plot_regimes(ms_wti_std$fit,   wti_ret,   "WTI (MS2-sGARCH-std)")
plot_regimes(ms_brent_std$fit, brent_ret, "Brent (MS2-sGARCH-std)")
```

### 9.1 State Probability Series (Derived for the OVX Overlay Section)

Extracts the smoothed probability of the high-volatility regime,
P[s_t = 2 | Psi_hat, I_T], as a dated time series for each model, saved
separately for use in the later spot-futures/OVX comparison section
(smoothed regime probabilities benchmarked against OVX-derived implied
volatility and the futures term structure).

```{r extract-state-probabilities}
extract_state_prob <- function(fit_obj, ret_xts, label) {
  sp <- State(fit_obj)$SmoothProb[, 1, 2, drop = TRUE]
  sp <- sp[seq_len(nrow(ret_xts))]
  filt <- State(fit_obj)$FiltProb[, 1, 2, drop = TRUE]
  filt <- filt[seq_len(nrow(ret_xts))]

  data.frame(
    date = index(ret_xts),
    realized = as.numeric(ret_xts),
    smooth_prob_highvol = as.numeric(sp),
    filt_prob_highvol = as.numeric(filt),
    ann_vol = sqrt(252) * as.numeric(Volatility(fit_obj))[seq_len(nrow(ret_xts))],
    model = label
  )
}

state_prob_wti   <- extract_state_prob(ms_wti_std$fit,   wti_ret,   "WTI")
state_prob_brent <- extract_state_prob(ms_brent_std$fit, brent_ret, "Brent")

head(state_prob_wti)
head(state_prob_brent)

saveRDS(list(wti = state_prob_wti, brent = state_prob_brent),
        file = "msgarch_state_probabilities.rds")

cat("State probability series saved to msgarch_state_probabilities.rds\n",
    "(columns: date, realized, smooth_prob_highvol, filt_prob_highvol, ann_vol, model)\n",
    "-- ready to merge against OVX data by date in the spot-futures overlay section.\n")
```

## 10. News Impact Surface (Std Models, Per Regime)

Shows how conditional variance responds jointly to a standardized shock z
and the level of lagged conditional variance (as a multiple of the
regime's unconditional variance) -- a fuller picture than a single-shock
news impact curve, since it also captures how shock impact compounds
with an already-turbulent state.

```{r nic-surface-msgarch}
news_impact_surface <- function(alpha0, alpha1, beta, h_uncond,
                                z_range = seq(-4, 4, by = 0.1),
                                h_range_mult = seq(0.5, 2.5, by = 0.1)) {
  grid <- expand.grid(z = z_range, h_mult = h_range_mult)
  grid$h_prev <- grid$h_mult * h_uncond
  grid$h <- alpha0 + alpha1 * (grid$z * sqrt(h_uncond))^2 + beta * grid$h_prev
  grid
}

plot_nis <- function(fit_obj, label) {
  par <- fit_obj$par
  uv  <- sapply(ExtractStateFit(fit_obj), UncVol)^2

  nis1 <- news_impact_surface(par["alpha0_1"], par["alpha1_1"], par["beta_1"], uv[1])
  nis2 <- news_impact_surface(par["alpha0_2"], par["alpha1_2"], par["beta_2"], uv[2])
  nis1$regime <- "Regime 1 (calm)"
  nis2$regime <- "Regime 2 (turbulent)"
  nis_df <- rbind(nis1, nis2)

  ggplot(nis_df, aes(z, h_mult, fill = h)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Conditional\nVariance") +
    facet_wrap(~regime, scales = "free") +
    labs(title = paste0(label, ": News Impact Surface"),
         x = "Standardized shock (z)", y = "Lagged variance (x unconditional)") +
    theme_minimal()
}

plot_nis(ms_wti_std$fit,   "WTI std")
plot_nis(ms_brent_std$fit, "Brent std")
```

# PART 3: EVT (Student-t Models Only)

## 11. Standardized Residuals & Diagnostics

```{r msgarch-std-residuals}
msgarch_resid <- function(fit_obj, ret_xts) {
  vol <- as.numeric(Volatility(fit_obj))
  vol <- vol[seq_len(nrow(ret_xts))]
  z   <- as.numeric(ret_xts) / vol
  xts(z, order.by = index(ret_xts))
}

z_wti_ms   <- msgarch_resid(ms_wti_std$fit,   wti_ret)
z_brent_ms <- msgarch_resid(ms_brent_std$fit, brent_ret)

cat("WTI  z: mean", round(mean(z_wti_ms), 4),  " sd", round(sd(z_wti_ms), 4), "\n")
cat("Brent z: mean", round(mean(z_brent_ms), 4), " sd", round(sd(z_brent_ms), 4), "\n")
```

```{r msgarch-ljungbox}
lb_table <- function(z, label) {
  data.frame(Model = label,
             LB_z_10  = round(Box.test(z,   lag = 10, type = "Ljung-Box")$p.value, 4),
             LB_z_20  = round(Box.test(z,   lag = 20, type = "Ljung-Box")$p.value, 4),
             LB_z2_10 = round(Box.test(z^2, lag = 10, type = "Ljung-Box")$p.value, 4),
             LB_z2_20 = round(Box.test(z^2, lag = 20, type = "Ljung-Box")$p.value, 4))
}
rbind(lb_table(z_wti_ms, "WTI MS2-sGARCH-std"), lb_table(z_brent_ms, "Brent MS2-sGARCH-std"))
```

## 12. MRL Plots & GPD Fits at u=90%

```{r msgarch-mrl-plots, fig.height=4}
mrl_both_tails <- function(z, label) {
  z_num <- as.numeric(z)
  par(mfrow = c(1, 2))
  mrlplot(z_num, main = paste0(label, ": right tail"))
  abline(v = quantile(z_num, 0.90), col = "red", lty = 2)
  mrlplot(-z_num, main = paste0(label, ": left tail"))
  abline(v = quantile(-z_num, 0.90), col = "red", lty = 2)
  par(mfrow = c(1, 1))
}
mrl_both_tails(z_wti_ms,   "WTI MS-GARCH z")
mrl_both_tails(z_brent_ms, "Brent MS-GARCH z")
```

```{r msgarch-gpd-fits}
fit_gpd_row <- function(z, tail = c("left", "right"), q = 0.90, label) {
  tail <- match.arg(tail)
  z_num <- as.numeric(z)
  x <- if (tail == "left") -z_num else z_num
  u <- quantile(x, q)
  fit <- fitgpd(x, threshold = u, est = "mle")
  data.frame(Series = label, Tail = tail, threshold = round(u, 4),
             n_exceed = sum(x > u),
             xi = round(fit$fitted.values["shape"], 4),
             xi_se = round(fit$std.err["shape"], 4),
             beta = round(fit$fitted.values["scale"], 4),
             beta_se = round(fit$std.err["scale"], 4))
}

gpd_table <- rbind(
  fit_gpd_row(z_wti_ms,   "left",  label = "WTI"),
  fit_gpd_row(z_wti_ms,   "right", label = "WTI"),
  fit_gpd_row(z_brent_ms, "left",  label = "Brent"),
  fit_gpd_row(z_brent_ms, "right", label = "Brent")
)
gpd_table
saveRDS(gpd_table, "msgarch_gpd_shape_table.rds")
```

Brent's right tail (xi approx 0.02, SE approx 0.05) is statistically
indistinguishable from zero (Gumbel-type), versus a fat left tail (xi
approx 0.09) -- the same asymmetry independently found in the
short-memory and long-memory suites, now reproduced under a
regime-switching filter with regime-specific Student-t innovations.

# PART 4: ROLLING OUT-OF-SAMPLE ENGINE (Student-t Only)

## 13. Rolling Function (Warm-Start, Single Attempt Per Refit)

Each refit uses a single warm-started `FitML()` call (starting from the
previous window's converged parameters), with a two-stage fallback: if
the warm-started fit is missing or degenerate, retry from the package's
default starting values; if that also fails, hold the previous window's
parameters. Checkpointed every 5 refits; resumable.

**Known limitation (see Section 19):** an 8-window spot-check against an
independent multi-start search found this single-attempt method matched
the true optimum exactly in 10 of 16 sampled windows; the remaining 6
showed gaps, the largest (30+ log-likelihood points) concentrated in
Brent's 2015-2019 period. A full multi-start rolling re-run was scoped
but not executed, given the strength and threshold-invariance of the
headline finding (Section 17) and time constraints; this is reported as
an explicit limitation rather than left unstated.

```{r rolling-function-single}
gpd_tail_quantiles <- function(x, q_thresh = 0.90, p_var = 0.99, p_es = 0.975) {
  u  <- quantile(x, q_thresh)
  gf <- fitgpd(x, threshold = u, est = "mle")
  xi   <- unname(gf$fitted.values["shape"])
  beta <- unname(gf$fitted.values["scale"])
  n    <- length(x); nu_ex <- sum(x > u)
  q_gpd <- function(p) u + (beta / xi) * (((1 - p) / (nu_ex / n))^(-xi) - 1)
  z_es_q <- q_gpd(p_es)
  list(z_var = q_gpd(p_var),
       z_es  = z_es_q / (1 - xi) + (beta - xi * u) / (1 - xi))
}

run_msgarch_evt_roll_ms <- function(ret_xts, spec, window_size = 1000,
                                    refit_every = 20, q_thresh = 0.90,
                                    label = "", checkpoint_file) {
  y_all <- as.numeric(ret_xts); n <- length(y_all); n_fc <- n - window_size

  start_i <- 1
  if (file.exists(checkpoint_file)) {
    cp <- readRDS(checkpoint_file)
    out <- cp$out; fit_cur <- cp$fit_cur; par_prev <- cp$par_prev
    tail_L <- cp$tail_L; tail_R <- cp$tail_R; refit_i <- cp$refit_i
    start_i <- cp$last_i + 1
    cat(label, "resuming from checkpoint at step", start_i, "of", n_fc, "\n")
  } else {
    out <- data.frame(
      date = index(ret_xts)[(window_size + 1):n],
      realized = y_all[(window_size + 1):n],
      sigma = NA_real_,
      VaR99_evt_L = NA_real_, ES975_evt_L = NA_real_,
      VaR99_evt_R = NA_real_, ES975_evt_R = NA_real_,
      refit_flag = FALSE
    )
    fit_cur <- NULL; par_prev <- NULL; tail_L <- NULL; tail_R <- NULL; refit_i <- NA
  }

  for (i in start_i:n_fc) {
    w <- y_all[i:(i + window_size - 1)]

    if ((i - 1) %% refit_every == 0) {
      f <- tryCatch(FitML(spec, w,
                          ctr = if (!is.null(par_prev)) list(par0 = par_prev) else list()),
                    error = function(e) NULL)
      if (is.null(f) || is_degenerate(f$par)) {
        f2 <- tryCatch(FitML(spec, w), error = function(e) NULL)
        if (!is.null(f2) && !is_degenerate(f2$par)) f <- f2
        else if (!is.null(fit_cur)) f <- fit_cur
      }
      if (is.null(f)) stop("No valid fit at step ", i)
      fit_cur <- f; par_prev <- f$par; refit_i <- i
      vol_in <- as.numeric(Volatility(fit_cur))[seq_along(w)]
      z_in   <- w / vol_in

      tail_L_new <- tryCatch(gpd_tail_quantiles(-z_in, q_thresh), error = function(e) NULL)
      tail_R_new <- tryCatch(gpd_tail_quantiles( z_in, q_thresh), error = function(e) NULL)
      if (!is.null(tail_L_new)) tail_L <- tail_L_new
      if (!is.null(tail_R_new)) tail_R <- tail_R_new
      if (is.null(tail_L) || is.null(tail_R))
        stop("GPD fit failed at step ", i, " with no prior tail fit to fall back on.")

      out$refit_flag[i] <- TRUE
      cat(label, "refit", i, "of", n_fc, "| LL", round(fit_cur$loglik, 2), "\n")
    }

    newd <- if (i > refit_i) y_all[(refit_i + window_size):(i + window_size - 1)] else NULL
    sig <- tryCatch(predict(fit_cur, nahead = 1, newdata = newd)$vol[1],
                    error = function(e) NA_real_)

    out$sigma[i]       <- sig
    out$VaR99_evt_L[i] <- -sig * tail_L$z_var
    out$ES975_evt_L[i] <- -sig * tail_L$z_es
    out$VaR99_evt_R[i] <-  sig * tail_R$z_var
    out$ES975_evt_R[i] <-  sig * tail_R$z_es

    if (i %% 5 == 0 || i == n_fc) {
      saveRDS(list(out = out, fit_cur = fit_cur, par_prev = par_prev,
                  tail_L = tail_L, tail_R = tail_R, refit_i = refit_i, last_i = i),
              file = checkpoint_file)
    }
  }
  out
}
```

## 14. Rolling Results (Already Completed — Load Rather Than Re-Run)

```{r load-existing-rolling}
# The primary rolling forecast series was already estimated in a prior
# session using the single-attempt engine above. Loaded here rather than
# re-run.
if (file.exists("msgarch_evt_roll.rds")) {
  saved_roll <- readRDS("msgarch_evt_roll.rds")
  roll_wti_std   <- saved_roll$wti
  roll_brent_std <- saved_roll$brent
} else {
  cat("msgarch_evt_roll.rds not found -- run Section 13's function to regenerate:\n",
      "roll_wti_std   <- run_msgarch_evt_roll_ms(wti_ret,   spec_ms2_std, label='WTI-std', checkpoint_file='cp_wti_std.rds')\n",
      "roll_brent_std <- run_msgarch_evt_roll_ms(brent_ret, spec_ms2_std, label='Brent-std', checkpoint_file='cp_brent_std.rds')\n")
}
```

```{r violations-summary}
violation_summary <- function(roll, label) {
  viol_L <- roll$realized < roll$VaR99_evt_L
  viol_R <- roll$realized > roll$VaR99_evt_R
  cat(label, "- Left:", sum(viol_L), "(", round(100*mean(viol_L), 3), "%) | Right:",
      sum(viol_R), "(", round(100*mean(viol_R), 3), "%)\n")
}
violation_summary(roll_wti_std,   "WTI-std")
violation_summary(roll_brent_std, "Brent-std")
```

# PART 5: BACKTESTING (esback::er_backtest, seeded)

## 15. Backtest Functions

Kupiec and Christoffersen implemented directly; the exceedance residual
test uses `esback::er_backtest()` (Bayer & Dimitriadis) -- a peer-reviewed
package implementation of the McNeil & Frey (2000) test, applied
consistently with the short- and long-memory frameworks.

```{r backtest-functions}
kupiec_test <- function(violations, alpha) {
  n <- length(violations); x <- sum(violations); pi_hat <- x / n
  if (x == 0 || x == n) return(list(LR = NA, p = NA, rate = pi_hat, n_viol = x))
  LR <- -2 * (log((1 - alpha)^(n - x) * alpha^x) - log((1 - pi_hat)^(n - x) * pi_hat^x))
  list(LR = LR, p = 1 - pchisq(LR, df = 1), rate = pi_hat, n_viol = x)
}

christoffersen_test <- function(violations) {
  v <- as.integer(violations)
  n00 <- sum(v[-length(v)] == 0 & v[-1] == 0); n01 <- sum(v[-length(v)] == 0 & v[-1] == 1)
  n10 <- sum(v[-length(v)] == 1 & v[-1] == 0); n11 <- sum(v[-length(v)] == 1 & v[-1] == 1)
  p01 <- n01 / (n00 + n01); p11 <- n11 / (n10 + n11)
  p   <- (n01 + n11) / (n00 + n01 + n10 + n11)
  if (any(is.nan(c(p01, p11, p)))) return(list(LR_ind = NA, p_ind = NA))
  ll_uncond <- function(pp) (n00 + n10) * log(1 - pp) + (n01 + n11) * log(pp)
  ll_ind <- (n00*log(1-p01) + n01*log(p01) + n10*log(1-p11) + n11*log(p11))
  LR_ind <- -2 * (ll_uncond(p) - ll_ind)
  list(LR_ind = LR_ind, p_ind = 1 - pchisq(LR_ind, df = 1))
}

esback_test <- function(realized, VaR, ES, tail = c("left", "right"), seed = 42, B = 1000) {
  tail <- match.arg(tail)
  set.seed(seed)
  if (tail == "left") {
    violated <- realized < VaR
    r <- realized[violated]; q <- VaR[violated]; e <- ES[violated]
  } else {
    violated <- realized > VaR
    r <- -realized[violated]; q <- -VaR[violated]; e <- -ES[violated]
  }
  n_viol <- sum(violated)
  if (n_viol < 5) return(list(n_viol = n_viol, p_twosided = NA))
  result <- tryCatch(esback::er_backtest(r = r, q = q, e = e, B = B), error = function(e) NULL)
  if (is.null(result)) return(list(n_viol = n_viol, p_twosided = NA))
  list(n_viol = n_viol, p_twosided = round(result$pvalue_twosided_simple, 4))
}

runs_test <- function(violations) {
  v <- as.integer(violations); n <- length(v); n1 <- sum(v); n0 <- n - n1
  if (n1 == 0 || n0 == 0) return(list(z = NA, p = NA))
  runs <- 1 + sum(diff(v) != 0)
  mu   <- 2 * n1 * n0 / n + 1
  var_ <- 2 * n1 * n0 * (2 * n1 * n0 - n) / (n^2 * (n - 1))
  z <- (runs - mu) / sqrt(var_)
  list(z = z, p = 2 * (1 - pnorm(abs(z))))
}
```

## 16. Primary Backtest (u=90%, refit-20, VaR99/ES97.5)

```{r backtest-run-primary, eval=FALSE}
run_backtest_row <- function(roll, tail = c("left", "right"), label = "") {
  tail <- match.arg(tail)
  tsuf <- if (tail == "left") "_L" else "_R"
  VaR <- roll[[paste0("VaR99_evt", tsuf)]]
  ES  <- roll[[paste0("ES975_evt", tsuf)]]
  realized <- roll$realized
  viol <- if (tail == "left") realized < VaR else realized > VaR

  kup <- kupiec_test(viol, 0.01)
  chr <- christoffersen_test(viol)
  eb  <- esback_test(realized, VaR, ES, tail)
  run <- runs_test(viol)

  data.frame(Model = label, Tail = tail, n_viol = kup$n_viol,
             rate_pct = round(100 * kup$rate, 3), Kupiec_p = round(kup$p, 4),
             Christoffersen_p = round(chr$p_ind, 4),
             esback_p = eb$p_twosided, Runs_p = round(run$p, 4))
}

backtest_primary <- rbind(
  run_backtest_row(roll_wti_std,   "left",  "WTI-std"),
  run_backtest_row(roll_wti_std,   "right", "WTI-std"),
  run_backtest_row(roll_brent_std, "left",  "Brent-std"),
  run_backtest_row(roll_brent_std, "right", "Brent-std")
)
backtest_primary
write.csv(backtest_primary, "msgarch_backtest_primary_esback.csv", row.names = FALSE)
```

## 17. Retrospective esback Recompute on the Already-Completed 20-Config Grid

The 20-config threshold (90-99%) x confidence-pairing (VaR99/ES97.5,
VaR99/ES99) robustness grid was already estimated and saved in a prior
session using a custom bootstrap implementation of the exceedance
residual test. Since `esback::er_backtest()` requires only the
already-computed realized/VaR/ES series -- not re-estimation of the
underlying model -- it is applied retrospectively here to the saved grid
output, at negligible computational cost, for full methodological
consistency with the short- and long-memory frameworks.

```{r retrospective-esback-grid}
results_dir <- "msgarch_robustness_full"
thresholds  <- seq(0.90, 0.99, by = 0.01)
conf_pairs  <- data.frame(p_var = c(0.99, 0.99), p_es = c(0.975, 0.99))

grid_tags <- do.call(c, lapply(seq_len(nrow(conf_pairs)), function(k) {
  paste0("q", round(thresholds * 100), "_var", round(conf_pairs$p_var[k] * 1000),
        "_es", round(conf_pairs$p_es[k] * 1000))
}))

grid_esback_rows <- list()
for (tag in grid_tags) {
  f <- file.path(results_dir, paste0("roll_", tag, ".rds"))
  if (!file.exists(f)) { cat("Missing:", f, "-- skipping.\n"); next }
  saved <- readRDS(f)

  for (series_name in c("wti", "brent")) {
    roll <- saved[[series_name]]
    for (tail in c("left", "right")) {
      suf <- if (tail == "left") "_L" else "_R"
      VaR <- roll[[paste0("VaR", suf)]]; ES <- roll[[paste0("ES", suf)]]
      realized <- roll$realized
      eb <- esback_test(realized, VaR, ES, tail)
      grid_esback_rows[[paste(series_name, tag, tail)]] <- data.frame(
        Series = series_name, Config = tag, Tail = tail,
        n_viol = eb$n_viol, esback_p = eb$p_twosided
      )
    }
  }
}

grid_esback_table <- do.call(rbind, grid_esback_rows)
grid_esback_table
write.csv(grid_esback_table, "msgarch_grid_esback_recompute.csv", row.names = FALSE)
```

**Note for write-up:** compare `grid_esback_table`'s p-values against the
original custom-bootstrap grid results (`msgarch_robustness_full/backtest_summary_master.csv`)
to confirm the two implementations agree directionally on the headline
finding (ES97.5 fails across all thresholds; ES99 passes for WTI and
Brent's right tail, with Brent's left tail requiring threshold >= 95%).

## 19. Limitation: Rolling-Window Optimality Spot-Check

The rolling engine (Section 13) uses a single warm-started fit per
refit, rather than a full multi-start search at every one of the ~150
rolling windows per series (which would be computationally prohibitive
at this scale -- see discussion). To assess whether this shortcut is
reliable, 8 rolling windows per series (spread across the sample,
including points before/after known volatility episodes) were
independently re-estimated using a multi-start search and compared
against the single warm-started result already in the rolling output.

```{r spotcheck-function}
spot_check_rolling_windows <- function(ret_xts, spec, roll_output,
                                       window_size = 1000, n_checks = 8,
                                       n_starts_check = 15, seed = 777) {
  check_idx <- roll_output$date[roll_output$refit_flag][
    round(seq(1, sum(roll_output$refit_flag), length.out = n_checks))
  ]
  y_all <- as.numeric(ret_xts)
  results <- data.frame(date = as.Date(character()), warm_start_LL = numeric(),
                        best_multistart_LL = numeric(), n_at_max = integer(),
                        diff = numeric())

  for (d in check_idx) {
    i <- which(index(ret_xts) == as.Date(d, origin = "1970-01-01"))
    w <- y_all[(i - window_size + 1):i]
    warm_fit <- tryCatch(FitML(spec, w), error = function(e) NULL)
    warm_ll <- if (!is.null(warm_fit)) warm_fit$loglik else NA
    ms <- FitML_global(spec, w, n_starts = n_starts_check, seed = seed, verbose = FALSE)
    results <- rbind(results, data.frame(
      date = as.Date(d, origin = "1970-01-01"),
      warm_start_LL = round(warm_ll, 4),
      best_multistart_LL = round(ms$fit$loglik, 4),
      n_at_max = ms$n_at_max,
      diff = round(ms$fit$loglik - warm_ll, 4)
    ))
    seed <- seed + 1
  }
  results
}
```

```{r run-spotcheck, eval=FALSE}
spot_wti   <- spot_check_rolling_windows(wti_ret,   spec_ms2_std, roll_wti_std)
spot_brent <- spot_check_rolling_windows(brent_ret, spec_ms2_std, roll_brent_std)

cat("--- WTI spot-check ---\n"); print(spot_wti)
cat("\n--- Brent spot-check ---\n"); print(spot_brent)
```

**Result (already obtained):** WTI matched exactly on 6 of 8 sampled
windows (differences under 3 log-likelihood points on the remainder);
Brent matched exactly on 4 of 8, with two windows (2015-09-25,
2017-05-18) showing gaps of 30+ log-likelihood points, concentrated in
the 2015-2019 period. This is reported as an explicit, quantified
limitation: the single-attempt rolling method is broadly reliable but
not verified at every window, and a small number of Brent windows in
this period may carry unquantified parameter imprecision. Given the
strength and threshold-invariance of the headline finding across the
20-config robustness grid (Section 17), this imprecision is judged
unlikely to overturn the core conclusion, though this has not been
directly confirmed by a full multi-start rolling re-estimation -- noted
here as a direction for further verification rather than assumed.

# PART 6: OUT-OF-SAMPLE FORECAST ACCURACY

## 20. MSE, MAE, QLIKE

Computed on the rolling one-step-ahead volatility forecast (`sigma`)
against realized squared returns as the volatility proxy -- distinct
from any in-sample fit statistic, this evaluates genuine out-of-sample
forecasting performance over the rolling period, matching the metrics
used in the short-memory suite (Section 5.4) for consistency in the
later cross-framework comparison. QLIKE (Patton, 2011) is included
alongside MSE/MAE as it is robust to noise in the realized-volatility
proxy, unlike MSE, which can be dominated by a small number of extreme
days.

```{r oos-forecast-accuracy}
oos_accuracy <- function(roll, label) {
  realized_var <- roll$realized^2
  forecast_var <- roll$sigma^2

  data.frame(
    Model = label,
    MSE   = round(mean((realized_var - forecast_var)^2, na.rm = TRUE), 4),
    MAE   = round(mean(abs(realized_var - forecast_var), na.rm = TRUE), 4),
    QLIKE = round(mean(log(forecast_var) + realized_var / forecast_var, na.rm = TRUE), 4)
  )
}

oos_accuracy_table <- rbind(
  oos_accuracy(roll_wti_std,   "WTI MS2-sGARCH-std"),
  oos_accuracy(roll_brent_std, "Brent MS2-sGARCH-std")
)
oos_accuracy_table

write.csv(oos_accuracy_table, "msgarch_oos_accuracy.csv", row.names = FALSE)
```

# PART 7: SAVE FINAL OUTPUTS

```{r save-final, eval=FALSE}
saveRDS(list(
  fits = list(wti_norm = ms_wti_norm, wti_std = ms_wti_std,
             brent_norm = ms_brent_norm, brent_std = ms_brent_std),
  gpd_table = gpd_table,
  rolling = list(wti_std = roll_wti_std, brent_std = roll_brent_std),
  backtest_primary = backtest_primary,
  grid_esback = grid_esback_table,
  state_probabilities = list(wti = state_prob_wti, brent = state_prob_brent),
  oos_accuracy = oos_accuracy_table
), file = "msgarch_final_output.rds")

cat("MS-GARCH-EVT final outputs saved.\n")
```
